---
title: 
output: 
  pdf_document:
    keep_tex: yes
header-includes:
  - \usepackage{float} 
  - \usepackage{longtable}
  - \pagenumbering{gobble}
  - \usepackage{caption}
  - \captionsetup[table]{labelformat=empty}
  - \usepackage{pdflscape}
  - \newcommand{\blandscape}{\begin{landscape}}
  - \newcommand{\elandscape}{\end{landscape}}
knit: (function(inputFile, encoding) {
  rmarkdown::render(inputFile, encoding = encoding, output_dir = "supplimentary") })
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
pacman::p_load(tidyverse, knitr, kableExtra, qwraps2, brms, broom.mixed, bnlearn, janitor)

# source("r-files/functions.R")
# source("r-files/methods_code.R")
# source("r-files/results_code.R")
load("all_data2x.RData")
```


\textbf{S1 Table. Summary of modifications to LESCA items.}

```{r}
 tab <- data.frame(stringsAsFactors=FALSE,
 Question = c(19, 21, 22, 27, 29, 36, 49, 61, 62),
 original = c("Beginning a new school experience  (beginning college, transferring college etc)",
                "Academic probation / ineligibility",
                "Being dismissed from dorm or other residence",
                "Financial problems concerning school",
                "Conflict with room-mate",
                "Suspended from team for non-academic reasons",
                "Being absent from school (classes) because of participation in sport",
                "Major change in level of academic performance (doing better or worse)",
                "Making career decisions (applying to graduate school,interviewing for jobs, etc)"),
 "Sportsstudents" = c("Beginning a new school experience (beginning university, transferring university etc)",
                       NA,
                       "Being dismissed from halls or other residence", 
                       NA, 
                       NA, 
                       NA,
                       "Being absent from university (classes) because of participation in sport",
                     NA,
                     "Making career decisions (applying for Masters degree, interviewing for jobs, etc)"),
 "non-student" = c("Beginning a new work experience",
                   "Removed",
                   "Being asked to vacate house/home",
                   "Financial problems",
                   "Conflict within household",
                   "Removed",
                   "Being absent from work because of participation in sport",
                   "Major change in level of performance at work (doing better or worse)",
                   "Making career decisions (applying for Masters degree, interviewing for jobs, etc)")) 

tab[is.na(tab)] <- "\\textit{No change}"

kable(
  tab %>%
    mutate(Sportsstudents = ifelse(str_detect(Sportsstudents, "-"), text_spec(Sportsstudents, align = "c"), Sportsstudents)),
  col.names = c("Q", "Original", "Students", "Non-students"),
  escape = F,
  caption = ""
) %>% 
  column_spec(1, width = c("0.3cm")) %>% 
   column_spec(2:4, width = ("4.8cm")) %>% 
  kable_styling(latex_options = "HOLD_position")
```

\newpage

\textbf{S2 Table. All arc strengths greater than 0.3 included in the first network.}

```{r results='asis'}
full <- str %>%
  filter(strength > 0.299 & direction != 0) %>%
  arrange(-strength) %>%
  `colnames<-`(str_replace_all(colnames(.), "_", "\\\\_")) %>%
    mutate(across(1:2, .fns = ~case_when(.x == "pi" ~ "Previous\ninjury",
                                       .x == "gender" ~ "Gender",
                                       .x == "clevel" ~ "Competitive\nlevel",
                                       .x == "hours" ~"Training hours",
                                       .x == "ind_team" ~ "Sport type",
                                       .x == "nlebase" ~ "Baseline\nNLE",
                                       .x == "stiffness_1" ~ "Stiffness_1",
                                       .x == "stiffness_2" ~ "Stiffness_2",
                                       .x == "NLE_1" ~ "Negative life\nevents_1",
                                       .x == "NLE_2" ~ "Negative life\nevents_2",
                                       .x == "RI_1" ~ "Reward\nInterest_1",
                                       .x == "RI_2" ~ "Reward\nInterest_2",
                                       .x == "BIS_1" ~ "Behavioural\nInhibition\nSystem_1",
                                       .x == "BIS_2" ~ "Behavioural\nInhibition\nSystem_2",
                                       .x == "HRV_1" ~ "Heart rate\nvariability_1",
                                       .x == "HRV_2" ~ "Heart rate\nvariability_2",
                                       .x == "balance_1" ~ "Balance_1",
                                       .x == "balance_2" ~ "Balance_2",
                                       .x == "FFFS_1" ~ "Fight-Flight-\nFreeze\nSystem_1",
                                       .x == "FFFS_2" ~ "Fight-Flight-\nFreeze\nSystem_2",
                                       .x == "injured_1" ~ "Injured_1",
                                       .x == "injured_2" ~ "Injured_2"
  ))) %>% 
   mutate_at(c("from", "to"), ~str_replace_all(., "_", "\\\\_")) %>% 
  dendroTools::round_df(2)

  kable(
    full,
    caption = "",
    longtable = T,
    escape = F,
    format = "latex"
  ) %>% 
  kable_styling(latex_options = c("HOLD_position" , "repeat_header"), repeat_header_text = "S3 Table continued")

```


\clearpage
\newpage

\textbf{S3 Table. All arc strengths greater than 0.3 included in the second network.}

```{r results='asis'}
strchange %>%
  filter(strength > 0.299 & direction != 0) %>%
  arrange(-strength) %>% 
  mutate(across(1:2, .fns = ~case_when(.x == "pi" ~ "Previous\ninjury",
                                       .x == "gender" ~ "Gender",
                                       .x == "clevel" ~ "Competitive\nlevel",
                                       .x == "hours" ~"Training\nhours",
                                       .x == "stiffness" ~ "Stiffness",
                                       .x == "nlec" ~ "Negative life\nevents",
                                       .x == "RI" ~ "Reward\ninterest",
                                       .x == "BIS" ~ "Behavioural\nInhibition\nSystem",
                                       .x == "HRV" ~ "Heart rate\nvariability",
                                       .x == "balance" ~ "Balance",
                                       .x == "FFFS" ~ "Fight-Flight-\nFreeze\nSystem",
                                       .x == "injured" ~ "Injured"))) %>%
  `colnames<-`(str_replace_all(colnames(.), "_", "\\\\_")) %>%
   mutate_at(c("from", "to"), ~str_replace_all(., "_", "\\\\_")) %>% 
  dendroTools::round_df(2) %>% 
  kable(
    caption = "",
    longtable = T,
    escape = F,
    format = "latex"
  ) %>% 
  kable_styling(latex_options = ("HOLD_position"))
```

\newpage

\textbf{S4 Table. Breakdown of injury count by sport and injury type (chronic or acute)}

```{r results = 'asis'}

injtable %>% 
  left_join(datanet %>% 
              select(id, sportg, time), c("id", "time")) %>% 
  tabyl(sportg, acu_over) %>% 
  adorn_percentages() %>% 
  adorn_pct_formatting() %>% 
  adorn_ns() %>% 
  select(sport = sportg, everything()) %>% 
  mutate(across(.fns = ~str_replace(.x, "%", "\\\\%"))) %>% 
  kable(
    caption = "",
    escape = F,
    format = "latex",
    align = "lcc"
  ) %>% 
  kable_styling(latex_options = ("HOLD_position"))
```

\newpage
\blandscape

\textbf{S5 Table. Breakdown of injury count by sport and injury location}

```{r results = 'asis'}

injtable %>% 
  left_join(datanet %>% 
              select(id, sportg, time), c("id", "time")) %>% 
  mutate(bodypart = ifelse(bodypart %in% c("foot_toe", "ankle_achilles"), "foot_ankle", bodypart)) %>% 
  tabyl(sportg, bodypart) %>% 
  adorn_totals() %>% 
  adorn_percentages() %>% 
  adorn_pct_formatting() %>% 
  adorn_ns() %>% 
  select(sport = sportg,  
         `foot / ankle` = foot_ankle, 
         `lower leg` = lowerleg, 
         knee,
         `upper leg` = upperleg, 
         hip, 
         torso, 
         `upper arm / shoulder` = upperarm_shoulder,
         `elbow / lower arm` = elbow_lowerarm_hand, 
         `head / neck` = head_neck) %>% 
  mutate(across(.fns = ~str_replace(.x, "%", "\\\\%"))) %>% 
  kable(
    caption = "",
    escape = F,
    format = "latex",
    align = "lccccccccc"
  ) %>% 
  kable_styling(latex_options = ("HOLD_position"))
```

\elandscape

\newpage

\textbf{S6 Table. Probabilities of injury conditional on the variables in the Markov blanket for injured\_1.}

```{r results='asis'}
set.seed(1001)

tab3 <- newprobtable(
  tallc,
  c("balance_1", "stiffness_1", "NLE_1", "hours", "clevel"),
  "injured_1",
  "injured",
  avg30
) %>%
  arrange(-prob) %>%
  select(outcome, everything()) %>%
  dendroTools::round_df(2) %>% 
  select(7, 2:6) %>% 
    rename(Probability = "prob",
         "Competitive level" = clevel,
         "Training hours" = hours,
         "Negative life events_1" = "NLE_1") %>% 
  `colnames<-`(str_replace_all(colnames(.), "_", "\\\\_")) %>% 
  mutate_all(~str_replace_all(., "_", "\\\\_"))

kable(
  tab3,
  caption = "",
    escape = F,
  align = "c"
) %>%
  kable_styling(latex_options = "HOLD_position")
```

\newpage

\textbf{S7 Table. Probabilities of injury conditional on the variables in the Markov blanket for injured\_2.}

```{r results='asis'}

set.seed(1002)

tab4 <- newprobtable(
  tallc,
  c("balance_2", "stiffness_2", "NLE_2", "FFFS_1", "HRV_2"),
  "injured_2",
  "injured",
  avg30
) %>%
  arrange(-prob) %>%
  select(outcome, everything()) %>%
  dendroTools::round_df(2) %>%
  select(7, 2:6) %>%
    rename(Probability = "prob",
         "Fight-Flight-Freeze System_1" = FFFS_1,
         "Negative life events_2 "= NLE_2,
         "Heart rate variability_2" = "HRV_2",
         Stiffness_2 = stiffness_2,
         Balance_2 = balance_2) %>%
  `colnames<-`(str_replace_all(colnames(.), "_", "\\\\_")) 


  kable(
    tab4,
    caption = "",
    escape = F,
    align = "c"
  ) %>%
  kable_styling(latex_options = "HOLD_position") %>% 
      column_spec(c(2,3),width = "2cm") %>% 
  column_spec(5, width = "2.8cm")

```


\newpage

\textbf{S8 Table. Highest and lowest probability of injury, conditional on the all variables in the Markov blanket for ``injured''.}

```{r results='asis'}
set.seed(1007)
testt <- expand.grid(list(
  c(-1:1),
  c(-1:1),
  c("Low", "High"),
  c("injury", "no injury")
))

query7 <- pmap_dfr(
  .l = testt,
  ~ tibble(
    prob = cpquery(
      fittedcont,
      event = (injured == "injured"),
      evidence = list(
        nlec = as.numeric(..1),
        stiffness = as.numeric(..2),
        hours = ..3,
        pi = ..4
      ),
      method = "lw"
    ),
    nle = ..1,
    stiffness = ..2,
    hours = ..3,
    pi = ..4
  )
) %>%
  arrange(-prob)

query7 %>%
  mutate_at(vars(nle, stiffness), ~ case_when(
    . == 1 ~ "+1SD",
    . == 0 ~ "mean",
    . == -1 ~ "-1SD"
  )) %>%
  select(prob, hours, pi, nle, stiffness) %>%
    rename(Probability = "prob",
         "Training hours" = hours,
         "Previous injury" = pi,
         "Negative life events" = nle,
         Stiffness = stiffness) %>%
  dendroTools::round_df(2) %>%
  kable(
    .,
    align = "ccccc"
  ) %>%
  kable_styling(latex_options = "HOLD_position") %>%
  row_spec(0,bold=TRUE)
```

