---
title: Supplimentary materials
output: 
  pdf_document:
    keep_tex: yes
header-includes:
  - \usepackage{float} 
  - \usepackage{longtable}
  - \pagenumbering{gobble}
  - \usepackage{caption}
  - \captionsetup[table]{labelformat=empty}
knit: (function(inputFile, encoding) {
  rmarkdown::render(inputFile, encoding = encoding, output_dir = "supplimentary") })
---

```{r include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
pacman::p_load(tidyverse, knitr, kableExtra, qwraps2, brms, broom.mixed, bnlearn)

source("r-files/functions.R")
source("r-files/methods_code.R")
source("r-files/results_code.R")
```


\textbf{S1 Table. Summary of modifications to LESCA items.}

```{r}
 tab <- data.frame(stringsAsFactors=FALSE,
 Question = c(19, 21, 22, 27, 29, 36, 49, 61, 62),
 original = c("Beginning a new school experience  (beginning college, transferring college etc)",
                "Academic probation / ineligibility",
                "Being dismissed from dorm or other residence",
                "Financial problems concerning school",
                "Conflict with room-mate",
                "Suspended from team for non-academic reasons",
                "Being absent from school (classes) because of participation in sport",
                "Major change in level of academic performance (doing better or worse)",
                "Making career decisions (applying to graduate school,interviewing for jobs, etc)"),
 "Sportsstudents" = c("Beginning a new school experience (beginning university, transferring university etc)",
                       NA,
                       "Being dismissed from halls or other residence", 
                       NA, 
                       NA, 
                       NA,
                       "Being absent from university (classes) because of participation in sport",
                     NA,
                     "Making career decisions (applying for Masters degree, interviewing for jobs, etc)"),
 "non-student" = c("Beginning a new work experience",
                   "Removed",
                   "Being asked to vacate house/home",
                   "Financial problems",
                   "Conflict within household",
                   "Removed",
                   "Being absent from work because of participation in sport",
                   "Major change in level of performance at work (doing better or worse)",
                   "Making career decisions (applying for Masters degree, interviewing for jobs, etc)")) 

tab[is.na(tab)] <- "\\textit{No change}"

kable(
  tab %>%
    mutate(Sportsstudents = ifelse(str_detect(Sportsstudents, "-"), text_spec(Sportsstudents, align = "c"), Sportsstudents)),
  col.names = c("Q", "Original", "Students", "Non-students"),
  escape = F,
  caption = ""
) %>% 
  column_spec(1, width = c("0.3cm")) %>% 
   column_spec(2:4, width = ("4.8cm")) %>% 
  kable_styling(latex_options = "HOLD_position")
```

\newpage

\textbf{S2 Table. All arc strengths greater than 0.3 included in the first network.}

```{r results='asis'}
full <- str %>%
  filter(strength > 0.299 & direction != 0) %>%
  arrange(-strength) %>%
  `colnames<-`(str_replace_all(colnames(.), "_", "\\\\_")) %>%
   mutate_at(c("from", "to"), ~str_replace_all(., "_", "\\\\_")) %>% 
  dendroTools::round_df(2) 

  kable(
    full,
    caption = "",
    longtable = T,
    escape = F,
    format = "latex"
  ) %>% 
  kable_styling(latex_options = c("HOLD_position" , "repeat_header"), repeat_header_text = "S3 Table continued")

```


\clearpage
\newpage

\textbf{S3 Table. All arc strengths greater than 0.3 included in the second network.}

```{r results='asis'}
strchange %>%
  filter(strength > 0.299 & direction != 0) %>%
  arrange(-strength) %>%
  `colnames<-`(str_replace_all(colnames(.), "_", "\\\\_")) %>%
   mutate_at(c("from", "to"), ~str_replace_all(., "_", "\\\\_")) %>% 
  dendroTools::round_df(2) %>% 
  kable(
    caption = "",
    longtable = T,
    escape = F,
    format = "latex"
  ) %>% 
  kable_styling(latex_options = ("HOLD_position"))
```

\newpage

\textbf{S4 Table. Probabilities of injury conditional on the variables in the Markov blanket for injured\_1.}

```{r results='asis'}
set.seed(1001)

tab3 <- newprobtable(
  tallc,
  c("balance_1", "stiffness_1", "nlelg_1", "hours", "clevel"),
  "injured_1",
  "injured",
  avg30
) %>%
  arrange(-prob) %>%
  select(outcome, everything()) %>%
  dendroTools::round_df(2) %>% 
  select(7, 2:6) %>% 
  `colnames<-`(str_replace_all(colnames(.), "_", "\\\\_")) %>% 
  mutate_all(~str_replace_all(., "_", "\\\\_"))

kable(
  tab3,
  caption = "",
    escape = F
) %>%
  kable_styling(latex_options = "HOLD_position")
```

\newpage

\textbf{S5 Table. Probabilities of injury conditional on the variables in the Markov blanket for injured\_2.}

```{r results='asis'}

set.seed(1002)

tab4 <- newprobtable(
  tallc,
  c("balance_2", "stiffness_2", "nlelg_2", "FFFS_1", "hrv_2"),
  "injured_2",
  "injured",
  avg30
) %>%
  arrange(-prob) %>%
  select(outcome, everything()) %>%
  dendroTools::round_df(2) %>%
  select(7, 2:6) %>%
  `colnames<-`(str_replace_all(colnames(.), "_", "\\\\_")) 


  kable(
    tab4,
    caption = "",
    escape = F
  ) %>%
  kable_styling(latex_options = "HOLD_position")

```


\newpage

\textbf{S6 Table. Highest and lowest probability of injury, conditional on the all variables in the Markov blanket for ``injured''.}

```{r results='asis'}
set.seed(1007)
testt <- expand.grid(list(
  c(-1:1),
  c(-1:1),
  c("Low", "High"),
  c("injury", "no injury")
))

query7 <- pmap_dfr(
  .l = testt,
  ~ tibble(
    prob = cpquery(
      fittedcont,
      event = (injured == "injured"),
      evidence = list(
        nlec = as.numeric(..1),
        stiffness = as.numeric(..2),
        hours = ..3,
        pi = ..4
      ),
      method = "lw"
    ),
    nle = ..1,
    stiffness = ..2,
    hours = ..3,
    pi = ..4
  )
) %>%
  arrange(-prob)

query7 %>%
  mutate_at(vars(nle, stiffness), ~ case_when(
    . == 1 ~ "+1SD",
    . == 0 ~ "mean",
    . == -1 ~ "-1SD"
  )) %>%
  select(prob, hours, pi, nle, stiffness) %>%
  rename(Probability = "prob") %>%
  dendroTools::round_df(2) %>%
  kable(
    .,
    align = "ccccc"
  ) %>%
  kable_styling(latex_options = "HOLD_position") %>%

  row_spec(0,bold=TRUE)
```

