---
title: "An interdisciplinary examination of stress and injury occurrence in athletes"
runtitle: "Short Title"
documentclass: frontiersHLTH  # or frontiersHLTH, or frontiersFPHY
author:
  - name: Dr Harry Fisher
    affiliation: '1'
    etal: Fisher # First author's last name. 
    email: harryfisher21@gmail.com # Indicates corresponding Author
    institution: Cardiff Metropolitan University
    street: Cyncoed Road
    zip: CF23 6XD
    country: United Kingdom
  - name: Dr Marianne Gittoes
    affiliation: '1'
  - name: Professor Lynne Evans
    affiliation: '1'
  - name: Miss Leah Bitchell
    affiliation: '1'  
  - name: Dr Richard Mullen  
    affiliation: '2'
  - name: Dr Marco Scutari
    affiliation: '3'    
affiliation:
  - id: '1'
    department: School of Sport and Health Science
    institution: Cardiff Metropolitan University
    city: Cardiff
    country: United Kingdom
  - id: '2' 
    department: Division of Sport, Health & Exercise Sciences
    institution: Brunel University
    city: London
    country: United Kingdom
  - id: '3' 
    institution: Istituto Dalle Molle di Studi sull'Intelligenza Artificiale (IDSIA) 
    city: Manno
    country: Switzerland
    
output: rticles::frontiers_article
csl: frontiers.csl
bibliography: ../library.bib
---

\begin{abstract}

This paper adopts an interdisciplinary approach to explore the relationship between psychosocial factors, physiological stress-related markers and occurrence of injury in athletes using a repeated measures design across a 2-year data collection period. At three data collection time-points, athletes completed measures of major life events, the reinforcement sensitivity theory personality questionnaire, muscle stiffness, heart rate variability and postural stability, and reported any injuries they had sustained since the last data collection. Two Bayesian networks were used to examine the relationships between variables and model the changes between data collection points in the study. Findings revealed muscle stiffness to have the strongest relationship with injury occurrence, with high levels of stiffness increasing the probability of sustaining an injury. Negative life events did not increase the probability of injury occurrence at any single time-point; however, when examining changes between time points, increases in negative life events did increase the probability of injury. In addition, the combination of increases in negative life events and muscle stiffness resulted in the greatest probability of sustaining an injury. Findings demonstrated the importance of both an interdisciplinary approach and a repeated measures design to furthering our understanding of the relationship between stress-related  markers and injury occurrence.

\tiny
 \keyFont{ \section{Keywords:} Sports Injury, Stress, Interdisciplinary, Bayesian Network} 

\end{abstract}


# Introduction {-}

Over the last four decades sport related injuries have received increased research attention [@Ivarsson2017].
This attention is unsurprising given the high incidence [@Rosa2014; @Sheu2016],
and undesirable physical and psychological effects of sports injuries [@Leddy1994; @Brewer2012].
To mitigate against both the increasing incidence and	undesirable consequences of injury, research  has identified several psychological [@Slimani2018],
anatomical [@Murphy2003],
biomechanical [@Neely1998; @Hughes2014]
and environmental [@Meeuwisse2007]
factors associated with sports injury occurrence.
Indeed, several models of injury causation have been proposed that highlight the multifactorial nature of injury occurrence [@Kumar2001; @Meeuwisse2007; @Wiese-Bjornstal2009],
of which one of the most widely cited was developed by Williams and Anderson [Fig \ref{fig:fig1}; @Andersen1988; @Williams1998].

\begin{figure}[!h]
\caption{{\bf Stress and injury model (Williams and Andersen, 1998).}}
\label{fig:fig1}
\end{figure}

Williams and Andersen's [@Williams1998] model proposed that when faced with a potentially stressful athletic situation, an athlete’s personality traits (e.g., hardiness, locus of control and competitive trait anxiety), history of stressors (e.g., major life events and previous injuries) and coping resources (e.g., general coping behaviours) will contribute to their response, either interactively or in isolation.
Central to the model is the stress response, which reflects the bi-directional relationship between athletes’ appraisal of, and response to, a stressful athletic situation. The model predicts that athletes who have a history of many stressors, personality traits that intensify the stress response and few coping resources, will exhibit greater attentional (e.g., peripheral narrowing) and/or physiological (e.g., increased muscle tension) responses that put these individuals at greater risk of injury.

Within Williams and Andersen’s [@Williams1998] model, major life events, a component of an athlete’s history of stressors, most consistently predicts injury occurrence [@Williams2007];
specifically, major life events with a negative, as opposed to positive, valence [@Passer1983a; @Maddison2005].
However, personality traits and coping resources have also been	found to predict injury, with for example, athletes more likely to sustain an injury if they have poor social support and psychological coping skills, and high trait anxiety and elevated competitive state anxiety; compared to athletes with the opposite profile. [@Smith1990; @Lavallee1996; @Ivarsson2010].
However, the amount of variance explained by the psychosocial factors proposed by the model has been modest, typically between 5 - 30% [@Galambos2005; @Ivarsson2010]; suggesting other factors are also likely to contribute to injury occurrence.

While the psychosocial factors proposed in @Williams1998 have received

Research that has explored the effects of the psychosocial factors in Williams and Andersen’s [@Williams1998] model, has failed to examine the mechanisms through which these factors are proposed to exert their effect.
To elaborate, the  model suggests that injuries are likely to occur through either increased physiological arousal resulting in increased muscle tension and reduced flexibility or attentional deficits caused by increased distractibility and peripheral narrowing.
However, to date, the research has largely focused on attentional deficits [@Andersen1999; @Rogers2005; @Wilkerson2012a; @Swanik2007].
For example, Andersen and Williams [@Andersen1999]
measured peripheral and central vision during high and low stress conditions and found athletes with high life event stress coupled with low social support had greater peripheral narrowing under stressful conditions compared to athletes with the opposing profile; these athletes went on to sustain an increased number of injuries during the following season.
Indeed, Rodgers and Landers [@Rogers2005] supported Andersen and Williams's [@Andersen1999] earlier findings reporting that peripheral narrowing under stress mediated 8.1% of the relationship between negative life events and injury.
However, few attempts have been made to explain the remaining variance between negative life events and athletic injury through the other proposed mechanisms, such as increased muscle tension and reduced motor control [cf. @Williams1998].

One possible reason for this oversight is the multifactorial nature of injury and the possible contribution of other non-psychological factors to the stress response
[@Meeuwisse2007; @Wiese-Bjornstal2009].
For example, a large body of research indicates that training-related stress is also likely to be related to the stress response and injury occurrence [@Lee2017; @Djaoui2017],
and may account for the unexplained variance from the psychological predictors of injury.
Appaneal and Perna [@Appaneal2014]
proposed the biopsychosocial model of stress athletic injury and health (BMSAIH) to serve as an extension to Williams and Andersen’s [@Williams1998] model and to address some of these issues.
To elaborate, the BMSAIH aimed to clarify the mediating pathways between the stress response and injury, consider other health outcomes and behavioural factors that impact sports participation,
and integrate the impact of training on athletes’ health [@Appaneal2014].
The central tenet of the BMSAIH is that psychosocial distress (e.g., negative life events) may act synergistically with training-related stress as a result of high-intensity and high-volume sports training, and “widen the window of susceptibility” [@Appaneal2014, p. 74] to a range of undesirable health outcomes including illness and injury.
Consequently, the BMSAIH provides a framework for future research to build on Williams and Andersen's [@Williams1998] model, by including other physiological markers of training-related stress, which together may provide greater insight into the injury process.

Although research supporting the BMSAIH has mainly focused on the relationship between hormonal responses to training and injury occurrence [@Perna1995; @Perna1997; @Perna2003], other research has identified additional markers of training-related stress that are associated with an increased risk of injury; for example,
heart rate variability [@Bellenger2016; @Williams2017],
postural stability [@Romero-Franco2014]
and muscle stiffness [@Pruyn2015].
Unfortunately, these markers are often studied in isolation without an assessment of the psychosocial factors that are known to contribute to injury, thereby limiting our understanding of how psychosocially and physiologically derived stress may contribute synergistically to injury occurrence.
Recently, Bittencourt et al. [@Bittencourt2016]
suggested that to better understand the multifactorial nature of sports injuries, research needs to move away from studying risk factors in isolation and instead adopt a complex systems approach to injury.
Such an approach posits that injury may arise from a complex “web of determinants” [@Bittencourt2016, p. 3],
where different factors interact in unpredictable and unplanned ways, but result in a global outcome pattern of either adaptation or injury.
Complex systems approaches have been used in health care to model the large number of risk factors  associated with different types of diseases [@Plsek2001];
however, very few studies have attempted to address sport injury occurrence using such an approach [@Hulme2018].

To summarise, despite offering a possible framework to build on the research stemming from Williams and Andersen's [@Williams1998] model, researchers have largely overlooked the potential to explore other physiological stress-related markers proposed by the BMSAIH, in addition to the already well-established psychological characteristics known to be related to injury [@Appaneal2014].
Furthermore, research has typically not captured changes in both psychosocial factors and stress-related physiological markers that may occur between initial measurement and injury occurrence.
Viewed through the lens of a complex systems approach the interaction between psychosocial sources of stress, stress-related physiological markers and injury occurrence may provide new insight into the injury process.
Therefore, the purpose of the current study was to examine the relationship between psychosocial factors, physiological stress-related markers and occurrence of injury in athletes over time.


# Results {-}

<!-- Comments should be HTML-style comments -->

## Subsection 1 {-}

You can use `R` chunks directly to plot graphs.  

```{r graph, echo = TRUE, message=FALSE, fig.show = "hide"}
x <- 0:100
set.seed(999)
y <- 2 * (x + rnorm(length(x), sd = 3) + 3)
plot(x, y)
```

## Subsection 2 {-}

Frontiers requires figures to be submitted individually, in the same order as
they are referred to in the manuscript. Figures will then be automatically
embedded at the bottom of the submitted manuscript. Kindly ensure that each
table and figure is mentioned in the text and in numerical order. Permission
must be obtained for use of copyrighted material from other sources (including
the web). Please note that it is compulsory to follow figure instructions.
Figures which are not according to the guidelines will cause substantial delay
during the production process.

# Discussion

# Disclosure/Conflict-of-Interest Statement {-}

<!--  
Frontiers follows the recommendations by the International Committee of Medical
Journal Editors (http://www.icmje.org/ethical_4conflicts.html) which require
that all financial, commercial or other relationships that might be perceived by
the academic community as representing a potential conflict of interest must be
disclosed. If no such relationship exists, authors will be asked to declare that
the research was conducted in the absence of any commercial or financial
relationships that could be construed as a potential conflict of interest. When
disclosing the potential conflict of interest, the authors need to address the
following points:

 - Did you or your institution at any time receive payment or services from a
   third party for any aspect of the submitted work?
 - Please declare financial relationships with entities that could be perceived
   to influence, or that give the appearance of potentially influencing, what
   you wrote in the submitted work.
 - Please declare patents and copyrights, whether pending, issued, licensed
   and/or receiving royalties relevant to the work.
 - Please state other relationships or activities that readers could perceive to
   have influenced, or that give the appearance of potentially influencing, what
   you wrote in the submitted work.
 -->

The authors declare that the research was conducted in the absence of any
commercial or financial relationships that could be construed as a potential
conflict of interest.

# Author Contributions {-}

<!--  

When determining authorship the following criteria should be observed:

 - Substantial contributions to the conception or design of the work; or the
   acquisition, analysis, or interpretation of data for the work; AND
 - Drafting the work or revising it critically for important intellectual
   content; AND
 - Final approval of the version to be published ; AND
 - Agreement to be accountable for all aspects of the work in ensuring that
   questions related to the accuracy or integrity of any part of the work are
   appropriately investigated and resolved.

Contributors who meet fewer than all 4 of the above criteria for authorship
should not be listed as authors, but they should be acknowledged.
(http://www.icmje.org/roles_a.html)

-->

The statement about the authors and contributors can be up to several sentences
long, describing the tasks of individual authors referred to by their initials
and should be included at the end of the manuscript before the References
section.


# Acknowledgments {-}

Funding:

# Supplemental Data 

Supplementary Material should be uploaded separately on submission, if there are
Supplementary Figures, please include the caption in the same file as the
figure. LaTeX Supplementary Material templates can be found in the Frontiers
LaTeX folder

# References

A reference list should be automatically created here. However it won't. Pandoc 
will place the list of references at the end of the document instead. There are 
no convenient solution for now to force Pandoc to do otherwise. The easiest way 
to get around this problem is to edit the LaTeX file created by Pandoc before 
compiling it again using the traditional LaTeX commands.

# Figures {-}

```{r, Figure-1, ref.label = "graph", results = "hide", echo = FALSE, message = FALSE, fig.height=4, fig.width=4, fig.align='center', fig.cap='Figure caption', out.width = "85mm", out.height = "85mm"}
# You can also refer to code chunks from above to place figures at the bottom.
```

